---
title: "Analysis hormonal pituitary data"
author: "Klaus Gröbner, Martino Bernasconi"
output: html_document
---

Data upload
```{r}

path<-"~/Documents/ETH/Eth_VI.SS21/BC_Epigenetic_Mechanisms_in_MentalHealth/Project_5_Materials-20210318/Data/"
setwd(path)
Pituitary_data<- read.csv("Pituitary_Data.csv", header= T)
meta_pit <- read.csv("metadata_pit.csv")
```
Libraries upload
```{r cars}
library(ggplot2)
library(limma)
library(dplyr)
```

## Prepare the data for analyses
Check whether the metadata ONLY contains the data we have in our data table. If there are additional samples stated in the metadata we removed them 

```{r pressure, echo=T}
identical(meta_pit$Name,Pituitary_data$Name)
```
We want the following structure of our pituitary data table:
Column names: pituitary hormone names (FSH, GH, …)
Row names: animal code names (A_001, A_002, …) 
Remove all columns, that don’t contain data (eg. the names column…)
Log-transform the pituitary data for analysis (use log() function)
```{r}
rownames(Pituitary_data)<- Pituitary_data$Name
Pituitary_data<-Pituitary_data[,-1]
Pituitary_data<- log(Pituitary_data)
head(Pituitary_data)
```
## Linear modelling:
Time point 1:
look at the time point 1 (T1), and compare between MSUS and CTRL only; for this we need to:
get the metadata for T1 only (try sub-setting based on T1)
get the pituitary data for T1 only 
```{r}
metaT1<-meta_pit[meta_pit$Timepoint=="T1",]
pituT1<-Pituitary_data[metaT1$Name,]
head(metaT1)
head(pituT1)
```
Then make a model matrix, where we model the hormonal levels on “Group” (MSUS/CTRL) 
model.matrix() allows us to crate create dummy variables (we have the controls as baseline)
```{r}
matrixGroup<-model.matrix(~Group, data = metaT1)
head(matrixGroup)
```
Fit the linear model
```{r}
fit <- eBayes(lmFit(t(pituT1), matrixGroup))
```
coefficients
```{r}
fit$coefficients
```
p-values
```{r}
fit$p.value
```
call the topTable() function on your object, to store all results of the model into a new variable
```{r}
resultsFit<-topTable(fit,n=Inf)
topTable(fit,n=Inf)

```
GH is significantly different and down-regulated in MSUS. In fact, the MSUS mice are generally smaller than Control.

Save table 
```{r}
write.csv(resultsFit, paste0(path, "Hormones_T1.csv"))
```
plotting 
```{r}
hormone<-c(rep("FSH",46),rep("GH",46),rep("LH",46),rep("Prolactin",46),rep("TSH",46))
level<- c(pituT1$FSH,pituT1$GH,pituT1$LH,pituT1$Prolactin,pituT1$TSH)
group<-c(rep(metaT1$Group,5))
df<-data.frame(hormone,level,group)

ggplot(df, aes(x=hormone, y=level, fill= group))+
  geom_boxplot()+
  geom_point(position=position_dodge(width=0.75), aes(group=group)) +
  annotate("text", x=2,y=5,label=paste0("P=",round(resultsFit$P.Value[1], digits=5)))+
  annotate("text", x=5,y=4,label=paste0("P=",round(resultsFit$P.Value[2], digits=5)))+
  annotate("text", x=3,y=7,label=paste0("P=",round(resultsFit$P.Value[3], digits=5)))+
  annotate("text", x=1,y=8,label=paste0("P=",round(resultsFit$P.Value[4], digits=5)))+
  annotate("text", x=4,y=7.5,label=paste0("P=",round(resultsFit$P.Value[5], digits=5)))+
  xlab("Hormone") +
  ylab("log(Level)") +
  labs(title="Hormone levels in Control and MSUS mice at T1") 
```

## Full model
reloading full dateset 
```{r}
rm(list=ls())
path<-"~/Documents/ETH/Eth_VI.SS21/BC_Epigenetic_Mechanisms_in_MentalHealth/Project_5_Materials-20210318/Data/"
Pituitary_data<- read.csv("Pituitary_Data.csv", header= T)
meta_pit <- read.csv("metadata_pit.csv")

```
Select  the parameters in the metadata that  are of biological importance for the linear model: 
Time point, group,  (ev individual), , housing is then removed because present only at T2 and T3.

```{r}
##define the levels
meta_pit$Housing<-relevel(as.factor(meta_pit$Housing),"SH")
levels(as.factor(meta_pit$Housing))
levels(as.factor(meta_pit$Timepoint))
levels(as.factor(meta_pit$Group))
levels(as.factor(meta_pit$Number))
```
Model matrix
```{r}
mm<-model.matrix(~Group*Timepoint*Housing,data = meta_pit)
head(mm)
```

## Fit the model 
sub-setting the model matrix
```{r}
mmTime<-mm[,c(1:4,6:8)]
rownames(Pituitary_data)<- Pituitary_data$Name
Pituitary_data<-Pituitary_data[,-1]
Pituitary_data<-log(Pituitary_data)
head(mmTime)
```
fit the model
```{r}
fitfull <- eBayes(lmFit(t(Pituitary_data), mmTime))
```
coefficients 
```{r}
fitfull$coefficients
```
p-values
```{r}
fitfull$p.value
```
toptable
```{r}
topTable(fitfull,n=Inf)
resultsfullTime<-topTable(fitfull,n=Inf)
resultsfullTime
```
saving stats for each coefficient 
```{r}
cols<-colnames(mmTime)
for (i in (1:length(cols))){
  res<-topTable(fitfull,coef=cols[i])
  write.csv(res, paste0(path,"results_half_model/","Hormones_Time",cols[i],".csv"))
  
}
```
## New model full model
subs-setting dataset for all without Housing and MSUSxHousing
```{r}
mmfull2<-mm[,-c(5,8)]
head(mmfull2)
```
fit the linear model 
```{r}
fitfull2 <- eBayes(lmFit(t(Pituitary_data), mmfull2))
```
coefficients
```{r}
fitfull2$coefficients
```
p-values
```{r}
fitfull2$p.value
```
toptable
```{r}
topTable(fitfull2,n=Inf)
resultsfull2<-topTable(fitfull2,n=Inf)
```
saving stats for each coefficient
```{r}
cols<-colnames(mmfull2)
for (i in (1:length(cols))){
  res2<-topTable(fitfull2,coef=cols[i])
  write.csv(res, paste0(path,"results_half_model/","Hormones_full_",cols[i],".csv"))
  }
```
GH is the only hormone significant in first full model
Box-plot for GH over time 
```{r}

combination<-c()
for (j in 1:137) {
 combination[j]<- paste0(meta_pit$Group[j], meta_pit$Housing[j])
}
dfGH<-data.frame(Pituitary_data$GH,meta_pit$Timepoint,meta_pit$Group,combination)


ggplot(dfGH, aes(y=Pituitary_data.GH, x=meta_pit.Timepoint, fill=combination))+
  geom_boxplot()+
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+    
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of GH at different time points")+
  labs(fill="groups")
```
```{r}
hormone2<-c(rep("FSH",137),rep("GH",137),rep("LH",137),rep("Prolactin",137),rep("TSH",137))
level2<- c(Pituitary_data$FSH,Pituitary_data$GH,Pituitary_data$LH,Pituitary_data$Prolactin,Pituitary_data$TSH)
dfMSUS<-data.frame(hormone2,level2,meta_pit$Group,meta_pit$Housing,meta_pit$Timepoint,combination)

```
Box-plot showing the level of hormones for the different group combinations at different time points
```{r}

ggplot(dfMSUS[meta_pit$Timepoint=="T1",], aes(x=hormone2, y=level2, fill=combination))+
  geom_boxplot()+
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+
  xlab("Hormones") +
  ylab("log(Level)") +
  labs(fill="groups")+
  labs(title="Hormones levels at time points T1") 
```
```{r}
ggplot(dfMSUS[meta_pit$Timepoint=="T2",], aes(x=hormone2, y=level2, fill=combination))+
  geom_boxplot()+
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+
  xlab("Hormones") +
  ylab("log(Level)") +
  labs(fill="groups")+
  labs(title="Hormones levels at time points T2") 
```
```{r}
ggplot(dfMSUS[meta_pit$Timepoint=="T3",], aes(x=hormone2, y=level2, fill=combination))+
  geom_boxplot()+
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+
  xlab("Hormones") +
  ylab("log(Level)") +
  labs(fill="groups")+
  labs(title="Hormones levels at time points T3") 
```
```{r}
dfProlactin<-data.frame(Pituitary_data$Prolactin,meta_pit$Timepoint,meta_pit$Group,combination)
ggplot(dfProlactin, aes(y=Pituitary_data.Prolactin, x=meta_pit.Timepoint, fill=combination))+
  geom_boxplot()+ 
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of Prolactin at different time points")+
  labs(fill="groups")
```
```{r}
dfLH<-data.frame(Pituitary_data$LH,meta_pit$Timepoint,meta_pit$Group,combination)
ggplot(dfLH, aes(y=Pituitary_data.LH, x=meta_pit.Timepoint, fill=combination))+
  geom_boxplot()+ 
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of LH at different time points")+
  labs(fill="groups")
```
```{r}
dfFSH<-data.frame(Pituitary_data$FSH,meta_pit$Timepoint,meta_pit$Group,combination)
ggplot(dfFSH, aes(y=Pituitary_data.FSH, x=meta_pit.Timepoint, fill=combination))+
  geom_boxplot()+ 
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of FSH at different time points")+
  labs(fill="groups")
```
```{r}
dfTSH<-data.frame(Pituitary_data$TSH,meta_pit$Timepoint,meta_pit$Group,combination)
ggplot(dfTSH, aes(y=Pituitary_data.TSH, x=meta_pit.Timepoint, fill=combination))+
  geom_boxplot()+ 
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of TSH at different time points")+
  labs(fill="groups")
```
Trend of FSH, TSH 
```{r}
ggplot(dfFSH[combination =="CONSH",], aes(y=Pituitary_data.FSH, x=meta_pit.Timepoint, fill=combination))+
  geom_boxplot()+ 
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of FSH at different time points")+
  labs(fill="groups")
```
```{r}
ggplot(dfTSH[combination =="CONSH",], aes(y=Pituitary_data.TSH, x=meta_pit.Timepoint, fill=combination))+
  geom_boxplot()+ 
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of TSH at different time points")+
  labs(fill="groups")
```
Plot prolactin only at time point T2 and T3 (only for presentation purpose)
```{r}
ggplot(dfProlactin[dfProlactin$meta_pit.Timepoint!="T1",], aes(y=Pituitary_data.Prolactin, x=meta_pit.Timepoint, fill=combination))+
  geom_boxplot()+ 
  geom_point(position=position_dodge(width=0.75), aes(group=combination))+
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of Prolactin at different time points")+
  labs(fill="groups")
```
why we can't use Housing EE in the model
```{r}
metaT1T2<-meta_pit[meta_pit$Timepoint!="T3",]
rownames(metaT1T2)<-metaT1T2$Name
pit12<-Pituitary_data[rownames(metaT1T2),]
summary(lm(pit12$GH~metaT1T2$Timepoint*metaT1T2$Housing))

```
we have no date for Housing EE at T1
