---
title: "Correlations pituitary and metabolics"
author: "Martino Bernasconi and Klaus Gröbner"
date: "1 4 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


load packages
```{r}
library(numbers)
library(limma)
library(ggplot2)
```


loading the data
```{r}
setwd("C:/Users/klaus/Documents/ETH/FS_2021/Epi_MH") 
path <-"C:/Users/klaus/Documents/ETH/FS_2021/Epi_MH/"
meta<-read.csv("metadata.csv")
mapped<-read.csv("mapped_data.csv")
mapped<-mapped[,-1]
meta_pit<-read.csv("metadata_pit.csv")
pit<-read.csv("Pituitary_Data.csv")
```

set up new metadata

```{r}
#create an empty list/array
pitID<-c()

#fill said list with all the IDs from the pituitary data that also have a sample in the metabolite data (works only because all the samples in the metabolite data have a matching sample in the hormone data)
for (i in (1:nrow(meta))){
  pitID<-c(pitID,meta_pit[which(meta_pit$Timepoint==meta$Timepoint[i]&meta_pit$Number==meta$Animal.No.[i]),][1])
}
pitID<-array(unlist(pitID))

#create the new metadata by adding the generated list as a column to the metadata of the metabolites
newmeta<-data.frame(pitID,meta)
```


Prepare the metabolic data
```{r}
#read in the metabolite data
mapped<-read.csv("mapped_data.csv")

#remove the first column because we don't need it
mapped<-mapped[,-1]

#remove duplicate metabolites
dups<-duplicated(mapped$Top.annotation.name)
sum(dups)
mapped<-mapped[!dups,]

#rename .1 columns by deleting the .1
colnames(mapped)<-sub(".1","",colnames(mapped),fixed=TRUE)
#head(mapped)

#create new matrix to store averaged data
uniqcols<-unique(colnames(mapped))
map<-matrix(data=NA,nrow=nrow(mapped),ncol = length(uniqcols))
colnames(map)<-uniqcols

#Fill said matrix up with the data
for (name in (uniqcols)){
  #temporarily store all the columns of one sample in a matrix
  temp<-as.matrix(mapped[,which(colnames(mapped)==name)])
  
  #if we don't have a duplicate we just take the sample
  if(ncol(temp)==1){
    map[,name]<-temp
  }
  
  #if we have duplicates we take the mean
  else{
     map[,name]<- rowMeans(temp)
  }
}

#set rownames to the metabolite names
map<-as.data.frame(map)
rownames(map)<-map$Top.annotation.name
map<-map[,-1]
#str(map)

#order the columns in map
map<-map[,order(colnames(map))]


#create the logtransformed matrix
temp<-matrix(log(as.numeric(as.matrix(map))),ncol = ncol(map))
rownames(temp)<-rownames(map)
colnames(temp)<-colnames(map)
logmap<-as.data.frame(temp)
```

select the correct data 
```{r}

cormap<-logmap


#choose all the metadata and pituitary data for which we have metabolite data
newmeta<-newmeta[newmeta$ï..Name %in% colnames(cormap),]
corpit<-pit[pit$ï..Name %in% newmeta$pitID,]

#check if the metadata and the actual data align
identical(colnames(cormap),newmeta$ï..Name)
identical(corpit$ï..Name,newmeta$pitID)
```
Extract animal numbers per group
```{r}
table(newmeta$Comb)
```

Run the correlation for each hormone metabolite pair

```{r}
#set up matrix for p-Value of correlation
pcor<-matrix(data = NA, ncol=ncol(corpit)-1, nrow=nrow(cormap))
colnames(pcor)<-colnames(corpit)[-1]
rownames(pcor)<-rownames(cormap)
pcor<-as.data.frame(pcor)

#fill the matrix up with the p-values
for (hormone in (colnames(pcor))){
  for (met in (rownames(pcor))){
    pcor[met,hormone]<-unlist(cor.test(as.numeric(cormap[met,]),corpit[,hormone])[3])
  }
}

```

Extract all the significant correlations
```{r}
FSHcorrs<-rownames(pcor[pcor[,1]<0.01,])
GHcorrs<-rownames(pcor[pcor[,2]<0.01,])
ProLaccorrs<-rownames(pcor[pcor[,3]<0.01,])
TSHcorrs<-rownames(pcor[pcor[,4]<0.01,])
LHcorrs<-rownames(pcor[pcor[,5]<0.01,])
```


run a linear model for each hormone-metabolite pair

```{r}
#generate new metadata by adding hormone columns to existing metadata
statmeta<-cbind(newmeta, corpit[,-1])

#generate the model matrix with Group hormones and group-hormone interaction
mm<-model.matrix(~Group*FSH+Group*GH+Group*Prolactin+Group*TSH+Group*LH, data = statmeta)
head(mm)

#set up matrix for the p-Values of the linear models
plin<-matrix(data = NA, ncol=ncol(corpit)-1, nrow=nrow(cormap))

colnames(plin)<-colnames(corpit)[-1]
rownames(plin)<-rownames(cormap)
plin<-as.data.frame(plin)
#fill the matrix up with the p-values
for (i in (1:ncol(plin))){
  for (met in (rownames(plin))){
    temp <- eBayes(lmFit(cormap[met,], mm[,c(1,2,i+2,i+7)]))
    plin[met,i]<-temp$p.value[4]
  }
}
```

Extract all the significant results
```{r}
FSHlin<-rownames(plin[plin[,1]<0.01,])
GHlin<-rownames(plin[plin[,2]<0.01,])
ProLaclin<-rownames(plin[plin[,3]<0.01,])
TSHlin<-rownames(plin[plin[,4]<0.01,])
LHlin<-rownames(plin[plin[,5]<0.01,])
```


Plotting Tyrosin, Pyridoxal, 4-pyridoxate over time for different combinations
```{r}
#transposing cormap and selecting the specific metabolites
t<-t(cormap)
tyrosin<-t[,"Tyrosine"]
Pyridoxal<-t[,"Pyridoxal"]
Pyridoxate<-t[,"4-Pyridoxate"]
Combination<-c()

#redefine all Housing at Timepoint 1 to SH
newmeta$Housing[which(newmeta$Timepoint=="T1")]<-"SH"

#Creating a new combined information column with Group and Housing
for (j in 1:105) {
 Combination[j]<- paste0(newmeta$Group[j],"/", newmeta$Housing[j])
  }

#Create a new dataframe with all the data we need
plotDf<-data.frame(Combination,newmeta$Timepoint,tyrosin,Pyridoxal,Pyridoxate)
plotDf<-setNames(plotDf, c("Combination","Timepoint","Tyrosine","Pyridoxal","4-Pyridoxate"))


# plot for tyrosine
ggplot(plotDf[], aes(y=Tyrosine, x=Timepoint, fill=Combination))+
  geom_boxplot()+ 
  geom_point(position=position_dodge(width=0.75), aes(group=Combination))+
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of Tyrosine at different time points")+
  labs(fill="groups")

```

plot for Pyridoxal
```{r}
ggplot(plotDf[], aes(y=Pyridoxal, x=Timepoint, fill=Combination))+
  geom_boxplot()+ 
  geom_point(position=position_dodge(width=0.75), aes(group=Combination))+
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of Pyridoxal at different time points")+
  labs(fill="groups")
```


plot for 4-Pyridoxate
```{r}
ggplot(plotDf[], aes(y=Pyridoxate, x=Timepoint, fill=Combination))+
  geom_boxplot()+ 
  geom_point(position=position_dodge(width=0.75), aes(group=Combination))+
  xlab("Time point") +
  ylab("log(Level)") +
  labs(title="Levels of Pyridoxal at different time points")+
  labs(fill="groups")
```






